---
output:
  html_document
bibliography: ref.bib
---

# Protein multiomics

```{r setup, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, error=FALSE)
```

## Motivation

So far, we've concerned ourselves with scRNA-seq data.
However, `r Biocpkg("scrapper")` also provides some functions to analyze single-cell proteomics data,
specifically from cellular indexing of transcriptomes and epitopes by sequencing (CITE-seq).
CITE-seq is a technique that quantifies both gene expression and the abundance of selected surface proteins in each cell simultaneously [@stoeckius2017simultaneous].
First, we create antibodies against the proteins of interest and conjugate them to synthetic RNA tags, i.e., antibody-derived tags (ADTs)^[
In theory, we could conjugate RNA to almost anything that sticks to a cell, e.g., cholersterol, peptides, small molecule drugs.
These tags are treated in the same way as ADTs, so we'll refer to all of them as ADTs for convenience.].
Cells are labelled with these antibodies and then processed with single-cell technologies like 10X Genomics.
For each cell, both ADTs and endogenous transcripts are reverse-transcribed into cDNA and sequenced.
This yields a set of counts for the ADTs, to quantify the abundance of each selected protein; and another set of counts for the genes, as in scRNA-seq.
We can then examine aspects of the proteome (e.g., post-translational modifications) and other cellular features that would normally be overlooked in transcriptomic studies.

Analysis of CITE-seq data is quite similar to that of scRNA-seq.
The RNA modality can simply be analyzed as if it came from an scRNA-seq experiment, and equivalent steps are performed for the ADT modality,
(e.g., quality control, normalization, PCA) with some tweaks to account for the differences between the ADT and RNA counts.
For example, the ADT data has fewer features as the proteins of interest were chosen by the reseacher,
and the coverage of each ADT is much deeper as the sequencing resources are concentrated into a smaller number of features.
At some point, we will also want to combine the ADT and RNA data so that information in both modalities are used in downstream steps like clustering.

## Quality control

As in the RNA-based analysis, we want to remove cells that have failed to capture/sequence the ADTs.
This involves similar QC metrics to those described in Chapter \@ref(quality-control), specifically:

- The number of features detected (i.e., with non-zero counts) in each cell.
  We expect non-zero counts for most ADTs in each cell, even if the corresponding protein target is not present on the cell surface.
  This is due to deeper sequencing coverage that detects up free-floating antibodies in the ambient solution or antibodies that are non-specifically bound to the cell membrance.
  An unusually low number of detected features is indicative of a failure in library preparation or sequencing.
- The sum of counts for isotype control (IgG) antibodies.
  IgG controls lack a specific target in the cell but otherwise have similar properties to the primary antibodies against proteins of interest.
  The coverage of these control ADTs serves as a measure of non-specific binding in each cell.
  A large sum for the controls is indicative of a problem with specificity, possibly even the formation of undesirable protein aggregates.

We demonstrate using a PBMC dataset from 10X Genomics [@zheng2017massively] that contains quantified abundances for a number of interesting surface proteins.

```{r}
library(DropletTestFiles)
path.pbmc <- getTestFile("tenx-3.0.0-pbmc_10k_protein_v3/1.0.0/filtered.tar.gz")
dir.pbmc <- tempfile()
untar(path.pbmc, exdir=dir.pbmc)

# Loading it in as a SingleCellExperiment object.
library(DropletUtils)
sce.pbmc <- read10xCounts(file.path(dir.pbmc, "filtered_feature_bc_matrix"))

# Splitting off the ADTs into an alternative experiment for separate
# processing, otherwise they'd be treated as genes.
sce.pbmc <- splitAltExps(sce.pbmc, rowData(sce.pbmc)$Type)
sce.pbmc

# Here, the "main" experiment contains the RNA data, while the alternative
# experiment contains the antibody data.
mainExpName(sce.pbmc)

sce.adt.pbmc <- altExp(sce.pbmc, "Antibody Capture")
sce.adt.pbmc

# Taking a sneak peak at the ADT counts.
counts(sce.adt.pbmc)[,1:10]
```

We compute each of the QC metrics described above from the ADT count matrix.
We also compute the sum of counts across all ADTs for each cell, but this is strictly for informational purposes only as it is not an effective QC metric.
The presence of a targeted protein can lead to a several-fold increase in the total ADT count, given the binary nature of most surface markers.
Removing cells with low total ADT counts could inadvertently eliminate cell types that do not express many - or indeed, any - of the selected protein targets.
Similarly, we prefer to use the sum of IgG counts instead of the proportion as the latter relies on the total count and is more affected by the biology.
For example, a cell that does not express any of the targets would have a lower total and thus a higher IgG proportion, making it unfairly susceptible to removal.

```{r}
library(scrapper)
is.igg.pbmc <- grep("^IgG", rownames(sce.adt.pbmc))
sce.qc.adt.pbmc <- quickAdtQc.se(sce.adt.pbmc, subsets=list(IgG=is.igg.pbmc))
summary(sce.qc.adt.pbmc$sum)
summary(sce.qc.adt.pbmc$detected)
summary(sce.qc.adt.pbmc$subset.sum.IgG)
```

The `quickAdtQc.se()` function computes thresholds using the outlier-based strategy described in Section \@ref(qc-outlier) (Figure \@ref(fig:qc-dist-adt-pbmc)).
We use a log-transformation for the number of detected features and the IgG sum to avoid negative thresholds and improve normality.
We also perform a minor adjustment to relax the threshold for the number of detected ADTs if the MAD is zero.

```{r qc-dist-adt-pbmc, fig.wide=TRUE, fig.cap="Distribution of ADT-based QC metrics in the PBMC dataset. Each point represents a cell, while dashed lines represent thresholds for each metric."}
qc.thresh.adt.pbmc <- metadata(sce.qc.adt.pbmc)$qc$thresholds
qc.thresh.adt.pbmc

library(scater)
gridExtra::grid.arrange(
    plotColData(sce.qc.adt.pbmc, y="detected") +
        geom_hline(yintercept=qc.thresh.adt.pbmc$detected, linetype="dashed", color="red") +
        ggtitle("Detected features"),
    plotColData(sce.qc.adt.pbmc, y="subset.sum.IgG") + 
        geom_hline(yintercept=qc.thresh.adt.pbmc$subset.sum["IgG"], linetype="dashed", color="red") +
        scale_y_log10() +
        ggtitle("IgG sum"),
    ncol=2
)
```

We then apply these thresholds to our metrics to identify high-quality cells.
If we wanted to use custom thresholds, we could modify our thresholds in the same manner as described in Section \@ref(qc-fixed).
Similarly, if our dataset contained multiple batches, we could use the same blocking approach as described in Section \@ref(qc-batch).

```{r}
summary(sce.qc.adt.pbmc$keep)
```

If we were only interested in the ADT data, we could subset our `SingleCellExperiment` with `qc.keep.adt.pbmc` and proceed to the next step.
However, we performed CITE-seq because we wanted to examine both protein abundance and gene expression for the same cell.
Thus, we need to apply quality control to the RNA counts as described in Chapter \@ref(quality-control).
We only keep cells that are considered to be of high quality for both their ADT and RNA modalities. 

```{r}
is.mito.pbmc <- grep("^MT-", rowData(sce.pbmc)$Symbol)
sce.qc.pbmc <- quickRnaQc.se(sce.pbmc, subsets=list(MT=is.mito.pbmc))

# Seeing how many cells pass both, one or neither QC filters. 
table(RNA=sce.qc.pbmc$keep, ADT=sce.qc.adt.pbmc$keep)

# Only keeping cells that pass both filters.
qc.keep.combined.pbmc <- sce.qc.pbmc$keep & sce.qc.adt.pbmc$keep
sce.qc.pbmc <- sce.qc.pbmc[,qc.keep.combined.pbmc]
sce.qc.adt.pbmc <- sce.qc.adt.pbmc[,qc.keep.combined.pbmc]
ncol(sce.qc.pbmc)
```

## Normalization

As with RNA, we performing scaling normalization to remove cell-specific biases due to differences in library preparation and sequencing efficiency (Chapter \@ref(normalization)).
Unfortunately, we can't just take the size factors for the RNA counts and re-use them for the ADTs.
The two modalities will be subject to different biases due to differences in biophysical properties between endogenous transcripts and the (much shorter) ADTs.
Some aspects of the library preparation and sequencing are also unique to each modality, providing more opportunities for differences in the biases.
So, instead, we need to compute ADT-specific size factors to normalize the ADT counts.

The simplest choice of size factor is to use the total sum of ADT counts, i.e., the library size for the ADTs.
Unfortunately, this is highly susceptible to composition biases caused by differences in protein abundance between cells.
Composition biases are much more pronounced in ADT data compared to RNA due to (i) the binary nature of target protein abundances,
where any increase in protein abundance manifests as a large increase to the total ADT count;
and (ii) the _a priori_ selection of interesting protein targets, which enriches for features that are more likely to be differentially abundant across the population.
These composition biases are strong enough to interfere with interpretation of fold-changes in protein abundance between clusters.

Instead, we use the geometric mean of all counts as the size factor for each cell [@stoeckius2017simultaneous].
This is based on the centered log-ratio (CLR) transformation for handling compositional data.
The geometric mean is a reasonable estimator of the scaling biases for large counts,
with the added benefit that it mitigates the effects of composition biases by dampening the impact of one or two highly abundant ADTs.
`r Biocpkg("scrapper")` implements a slightly more accurate variant of this approach named ["CLRm1"](https://github.com/libscran/clrm1/),
which accounts for the effect of adding a pseudo-count before computing the mean (which is required to avoid a mean of zero).
We center the size factors to ensure that the scaling normalization preserves the magnitude of the original counts,
and we compute log-normalized abundance values for ADTs as described in Section \@ref(norm-transformation).

```{r}
sce.norm.adt.pbmc <- normalizeAdtCounts.se(sce.qc.adt.pbmc)
summary(sce.norm.adt.pbmc$sizeFactor)
```

We observe some deviation between the CLRm1 size factors and their library size-derived counterparts (Figure \@ref(fig:cite-sf)). 
This is consistent with the presence of strong composition biases in the latter that are dampened in the former.
Of course, the geometric mean is not foolproof and will progressively become less accurate with more upregulated ADTs in each cell.
It is also more sensitive to noise at low counts, though this should be less problematic for ADT data due to its deeper sequencing coverage compared to RNA.

```{r cite-sf, fig.cap="Comparison between the CLRm1 size factors and the library size-derived factors for the ADT modality of the PBMC dataset."}
lib.sf.adt.pbmc <- centerSizeFactors(sce.norm.adt.pbmc$sum)
plot(sce.norm.adt.pbmc$sizeFactor, lib.sf.adt.pbmc, log="xy", pch=16, cex=0.5)
```

## Feature selection and PCA

Feature selection for ADTs is generally unnecessary as it was already performed during the design of the antibody panel.
The manual choice of target proteins means that all ADTs already correspond to "interesting" features.
In addition, there is little scope for further filtering when the number of ADTs is low.
Here, we have fewer than 20 ADTs, and even for the larger datasets, the panel will usually have less than 200 features.
These are small numbers compared to our previous selections of 1000-5000 HVGs in Chapter \@ref(feature-selection).

We might consider removing the IgG controls as we know that they will not be biologically interesting.
This probably won't make much difference as the controls are unlikely to exhibit strong variation that might intefere with downstream steps.
But it probably won't hurt either, so we might as well do it:

```{r}
selected.adt.pbmc <- !grepl("^IgG", rownames(sce.norm.adt.pbmc))
rowData(sce.norm.adt.pbmc)$of.interest <- selected.adt.pbmc
summary(selected.adt.pbmc)
```

We also perform a PCA on the ADT log-abundance matrix as described in Chapter \@ref(principal-components-analysis). 
This is mostly useful for datasets with larger panels to compact the data from ~200 ADTs to 10-20 PCs.
For smaller datasets, PCA is unnecessary as the number of ADTs is comparable to the typical number of PCs.
Regardless, it doesn't hurt to run a PCA in such cases -
if the number of ADTs is lower than the requested number of PCs, the PC scores will simply be a rotation of the log-abundance data.

```{r}
sce.pca.adt.pbmc <- runPca.se(
    sce.norm.adt.pbmc,
    features=selected.adt.pbmc,
    number=20
)
dim(reducedDim(sce.pca.adt.pbmc, "PCA"))
```

If we don't want to run a PCA, we could instead use the log-normalized abundance matrix directly in downstream analyses.

```{r}
# Transpose to make it look like a reducedDim entry, so that we could plug it
# into downstream algorithms by just setting reddim.type=.
norm.adt.pbmc <- t(assay(sce.norm.adt.pbmc, "logcounts")[selected.adt.pbmc,])
reducedDim(sce.pca.adt.pbmc, "selected") <- as.matrix(norm.adt.pbmc)

# For example...
sce.kmeans.adt.pbmc <- clusterKmeans.se(sce.pca.adt.pbmc, k=10, reddim.type="selected")
summary(sce.kmeans.adt.pbmc$clusters)
```

## The rest of the analysis {#cite-rest}

Once we have the PCs, we can use them for clustering and visualization in the same manner as described in Chapters \@ref(visualization) and \@ref(clustering). 
This summarizes the heterogeneity specific to the ADT modality (Figure \@ref(fig:cite-tsne)).

```{r cite-tsne, fig.cap="$t$-SNE plot generated from the log-normalized abundance of each ADT in the PBMC dataset. Each point is a cell and is colored according to its assigned cluster."}
sce.nn.adt.pbmc <- runAllNeighborSteps.se(sce.pca.adt.pbmc)
table(sce.nn.adt.pbmc$clusters)

library(scater)
plotReducedDim(sce.nn.adt.pbmc, "TSNE", colour_by="clusters")
```

We then identify markers from the log-abundance matrix, as described in Chapter \@ref(marker-detection).
For the top ADTs, we usually observe very large effect sizes due to the binary nature of surface targets.
However, there are also strong composition biases in this data so some caution is required when interpreting the smaller log-fold changes.

```{r}
markers.adt.pbmc <- scoreMarkers.se(sce.nn.adt.pbmc, sce.nn.adt.pbmc$clusters)
previewMarkers(markers.adt.pbmc[["1"]]) # Looking at the top marker tags for cluster 1.
```

We can also use the ADT-derived clusters to identify marker genes from the log-expression matrix for the RNA modality.
This is analogous to performing FACS to isolate cell types before differential expression analyses with bulk RNA-seq.

```{r}
# Computing log-normalized expression values from the RNA counts.
sce.norm.pbmc <- normalizeRnaCounts.se(sce.qc.pbmc)

# Computing markers for RNA data but using the ADT-derived clusters!
markers.adt2rna.pbmc <- scoreMarkers.se(sce.norm.pbmc, sce.nn.adt.pbmc$clusters, extra.columns="Symbol")

# Now looking at the top marker genes for cluster 1.
previewMarkers(markers.adt2rna.pbmc[["1"]], pre.columns="Symbol")
```

Conversely, we could use the RNA-derived clusters to score the ADTs.
This is most relevant when the ADTs represent some kind of functional readout (e.g., binding activity) instead of cell type identity.

## Combining modalities

A possibly more effective use of our data would consider heterogeneity in both modalities simultaneously.
In other words, the ADT and RNA data are combined in some manner prior to clustering and visualisations.
This ensures that any unique variation in either modality will be captured in the cluster definitions.
For example, if the antibody panel captures transient post-translation modifications like phosphorylation, this will not show up in the RNA data;
conversely, biological processes without a surface target will not be represented in the ADT data.
To demonstrate, let's continue the analysis of the RNA modality of our PBMC dataset:

```{r}
sce.var.pbmc <- chooseRnaHvgs.se(sce.norm.pbmc)
sce.pca.pbmc <- runPca.se(sce.var.pbmc, features=rowData(sce.var.pbmc)$hvg, number=20)
ncol(reducedDim(sce.pca.pbmc))
```

Possibly the simplest method to combine modalities involves literally combining the matrices of ADT- and RNA-derived PC scores.
(Or if no PCA was performed for the ADTs, the log-abundance matrix can be used instead.)
The combined matrix contains both sets of PCs, ensuring that information from both modalities will be used, e.g., when computing distances and finding neighbors.
However, naively combining the two matrices is not ideal as the number of genes is typically at several orders of magnitude greater than the number of ADTs.
This would cause the RNA modality to dominate the variance in the combined matrix, effectively sidelining any contributions from the ADT modality.

Instead, we scale the modalities to balance their contributions to the combined matrix with the `scaleByNeighbors.se()` function.
For each modality, we compute the median distance from each cell to its $k$ nearest neighbor, which we treat as a proxy for the uninteresting variation within subpopulations. 
Each matrix of PCs is then scaled according to its median distance, equalizing the magnitude of uninteresting variation across modalities.
This ensures that high baseline variation in one modality will not drown out interesting biological variation in another modality in the combined matrix.
We do not force modalities to have the same total variance, as the biological variation may genuinely differ across modalities.

```{r}
# We put our ADT experiment back inside the parent object so that
# scaleByNeighbors.se can see both sets of PCs at once.
altExp(sce.pca.pbmc, "Antibody Capture") <- sce.pca.adt.pbmc

sce.combined.pbmc <- scaleByNeighbors.se(
    sce.pca.pbmc,
    main.reddims="PCA",
    altexp.reddims=c(`Antibody Capture`="PCA")
)
dim(reducedDim(sce.combined.pbmc, "combined"))

# Scaling applied to the main experiment, i.e., the RNA.
metadata(sce.combined.pbmc)$combined$main.scaling

# Scaling applied to the alternative experiment, i.e., the ADTs.
metadata(sce.combined.pbmc)$combined$altexp.scaling
```

The combined matrix is convenient as it can be directly used in clustering or visualization (Figure \@ref(fig:cite-combined)) as if we were dealing with our usual matrix of PCs.

```{r cite-combined, fig.cap="$t$-SnE plot of the PBMC data generated from combined ADT and RNA PCs. Each point is a cell and is colored according to the assigned cluster."}
sce.nn.combined.pbmc <- runAllNeighborSteps.se(sce.combined.pbmc, reddim.type="combined")
table(sce.nn.combined.pbmc$clusters)
plotReducedDim(sce.nn.combined.pbmc, "TSNE", colour_by="clusters")
```

In practice, the RNA and ADT modalities are often strongly correlated when the antibody panel targets cell type-related proteins.
Using a combined matrix does not offer much benefit in these cases - in fact, we would say that a well-designed panel is more than enough for cell type identification^[
At least for blood, which has been FACS'd to death.], without any help from gene expression at all.
Combining modalities may even be detrimental if one of the modalities has little biological variation, e.g., if no antibodies are bound, the ADT matrix will only be contributing noise.
So, what should we do?
Well, our usual advice for single-cell analysis applies, a.k.a., see if we get interesting results and try something else if we don't.

## Session information {-}

```{r}
sessionInfo()
``` 
