---
output:
  html_document
bibliography: ref.bib
---

# Marker gene detection {#marker-detection}

```{r, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, error=FALSE)
```

## Motivation

Now that we've got our clustering results from Chapter \@ref(clustering), our next step is to identify the genes that drive separation between clusters.
Genes that are strongly upregulated in a particular cluster are called "markers" as they define the corresponding cell type/state relative to other cells in the population.
By examining the annotated functions of the marker genes, we can assign biological meaning to each cluster.
In the simplest case, if we know that certain genes are upregulated in a particular cell type, clusters with increased expression of the former can be identified as the latter.
More subtle cell states (e.g., activation status, stress) can also be determined based on the behavior of genes in the affected pathways.
To detect markers, we simply test for differential expression (DE) between clusters and examine the top DE genes from each comparison.

## Scoring marker genes

### Comparing pairs of clusters 

We quantify differential expression between each pair of clusters with different effect sizes (Section \@ref(marker-effect-sizes)) 
and then summarize the effect sizes across comparisons for each cluster (Section \@ref(marker-effect-summaries)).
These effect size summaries can then be used to obtain a ranking of potential marker genes for each cluster.
To illustrate, let's pull out our old friend, the PBMC dataset from 10X Genomics [@zheng2017massively].

```{r}
# Loading in raw data from the 10X output files.
library(DropletTestFiles)
raw.path.10x <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/filtered.tar.gz")
dir.path.10x <- file.path(tempdir(), "pbmc4k")
untar(raw.path.10x, exdir=dir.path.10x)

library(DropletUtils)
fname.10x <- file.path(dir.path.10x, "filtered_gene_bc_matrices/GRCh38")
sce.10x <- read10xCounts(fname.10x, col.names=TRUE)

# Applying our default QC with outlier-based thresholds.
library(scrapper)
is.mito.10x <- grepl("^MT-", rowData(sce.10x)$Symbol)
sce.qc.10x <- quickRnaQc.se(sce.10x, subsets=list(MT=is.mito.10x))
sce.qc.10x <- sce.qc.10x[,sce.qc.10x$keep]

# Computing log-normalized expression values.
sce.norm.10x <- normalizeRnaCounts.se(sce.qc.10x, size.factors=sce.qc.10x$sum)

# We now choose the top HVGs.
sce.var.10x <- chooseRnaHvgs.se(sce.norm.10x, more.choose.args=list(top=4000))

# Running the PCA on the HVG submatrix.
sce.pca.10x <- runPca.se(sce.var.10x, features=rowData(sce.var.10x)$hvg, number=25)

# Doing some graph-based clustering, t-SNEs, etc.
sce.nn.10x <- runAllNeighborSteps.se(sce.pca.10x)
sce.nn.10x
```

Given the clustering and the log-expression values, the `scoreMarkers.se()` function returns a data frame of marker statistics for each cluster.
Each data frame contains the mean log-expression and the proportion of cells with detected (i.e., non-zero) expression in a particular cluster.
It also contains multiple columns representing effect size summaries, where each column is named as `<effect size>.<summary type>`,
e.g., `cohens.d.mean` contains the mean of the Cohen's $d$ across all comparisons involving that cluster.
Genes with larger `cohens.d.mean` values exhibit stronger upregulation in the current cluster compared to the other clusters.
By default, `scoreMarkers.se()` orders the rows on `cohens.d.mean` to provide one possible ranking of markers for each cluster.

```{r}
markers.10x <- scoreMarkers.se(sce.nn.10x, sce.nn.10x$clusters, extra.columns="Symbol")
names(markers.10x)

# Examining the statistics for cluster 1.
chosen.cluster <- "1"
chosen.markers.10x <- markers.10x[[chosen.cluster]]
head(chosen.markers.10x)

# A more concise overview.
previewMarkers(chosen.markers.10x, pre.columns='Symbol')
```

We examine the top genes for some annotated biological function or cell type specificity that could be used to identify cluster `r chosen.cluster`^[
As of time of writing, the top genes were _LYZ_, _S100A8_ and _MNDA_, suggesting that this cluster contains monocytes.
But I'm not an immunologist, so take that with a grain of salt.].
Usually the first 10-20 genes are sufficient to assign some biological meaning to the cluster,
though we can always perform a more detailed characterization by considering additional genes from the ranking.
We can also visualize the distribution of their expression values in each cluster (Figure \@ref(fig:marker-violins-pbmc)),
where the top markers should be upregulated in cluster `r chosen.cluster` compared to most of the other clusters.

```{r marker-violins-pbmc, fig.wide=TRUE, fig.asp=2, fig.cap=paste0("Distribution of log-expression values for the top marker genes of cluster ", chosen.cluster, " in the PBMC dataset.")}
library(scater)
# Displaying symbols instead of Ensembl IDs for easier interpretation. 
plotExpression(
    sce.nn.10x,
    x="clusters",
    colour_by="clusters", # for some verisimilitude
    features=chosen.markers.10x$Symbol[1:6],
    swap_rownames="Symbol"
)
```

Of course, the default `cohens.d.mean` is just one of many possible choices for ranking potential marker genes.
Different effect sizes or summary statistics can yield alternative rankings that may be more or less useful.

### Choice of effect size {#marker-effect-sizes}

For each pairwise comparison, we compute several effect sizes to quantify the magnitude of differential expression.
The choice of effect size influences the types of markers that are prioritized in rankings based on that effect size.
To demonstrate, we'll look at the different rankings obtained for cluster `r chosen.cluster` with each effect size.
(For consistency and simplicity, we will use the `.mean` summary in each example.)

Cohen's $d$ is defined as the difference in the mean between groups divided by the average standard deviation across groups.
In other words, it is the number of standard deviations that separate the means of the two groups.
When applied to log-expression values, Cohen's $d$ can be interpreted as a standardized log-fold change.
Positive values indicate that the gene is upregulated in our cluster of interest, 
negative values indicate downregulation and values close to zero indicate that there is little difference.
Cohen's $d$ is roughly analogous to the $t$-statistic in a two-sample $t$-test.

```{r}
previewMarkers(chosen.markers.10x, pre.columns='Symbol', order.by="cohens.d.mean")
```

The area under the curve (AUC) is the probability that a randomly chosen observation from our cluster of interest is greater than a randomly chosen observation from the other cluster.
A value of 1 corresponds to upregulation, where all values of our cluster of interest are greater than any value from the other cluster;
a value of 0.5 means that there is no difference in the location of the distributions;
and a value of 0 corresponds to downregulation.
The AUC is closely related to the $U$ statistic in the Wilcoxon ranked sum test (a.k.a., Mann-Whitney U-test).
Both the AUC and Cohen's $d$ tend to detect similar markers.
The AUC is more robust to outliers but less sensitive to the magnitude of the differences between clusters,
i.e., a greater difference between clusters will usually result in a larger Cohen's $d$ but may not change the AUC much if it's already close to 0 or 1.

```{r}
previewMarkers(chosen.markers.10x, pre.columns='Symbol', order.by="auc.mean")
```

The "delta-detected" is the difference in the proportion of cells with detected (non-zero) expression between two clusters.
A value of 1 indicates that all cells in the cluster of interest express a gene, while all cells in the other cluster do not;
a value of zero indicates that there is no difference in the proportion;
and a value of -1 indicates that expression is only found in the cells of the other cluster. 
Rankings based on the delta-detected value will prioritize genes that are near-silent in the other cluster (Figure \@ref(fig:marker-detected-pbmc)).
When available, these genes are often very effective markers as they are only expressed in our cluster of interest.
However, it is also possible that strong markers will not have a large delta-detected value, e.g., because they are expressed at a low constitutive level in the other cluster.

```{r marker-detected-pbmc, fig.asp=2, fig.wide=TRUE, fig.cap=paste0("Distribution of log-expression values for the top marker genes of cluster ", chosen.cluster, " in the PBMC dataset, ranked by the mean delta-detected.")}
top.detected.10x <- previewMarkers(chosen.markers.10x, pre.columns='Symbol', order.by="delta.detected.mean")
top.detected.10x

plotExpression(
    sce.nn.10x,
    features=top.detected.10x$Symbol[1:6],
    swap_rownames="Symbol",
    x="clusters",
    colour_by="clusters"
)
```

Finally, the "delta-mean" is the difference in the mean between two clusters.
When computed on the log-normalized expression values, this is simply a fancy name for the log-fold change between clusters.
In most cases, Cohen's $d$ or the AUC are better choices as they account for the variance within each cluster.
Nonetheless, the log-fold changes are still useful as their values are easier to interpret.
They can also help to diagnose pathological situations where large Cohen's $d$ or AUC values are driven by small variances instead of large differences between clusters.

```{r}
# Same as the 'lfc' column.
previewMarkers(chosen.markers.10x, pre.columns='Symbol', order.by="delta.mean.mean")
```

Keep in mind that effect sizes are defined relative to other clusters in the same dataset.
Biologically meaningful genes will not be detected as markers if they are expressed uniformly throughout the population,
e.g., T cell markers will not be detected if only T cells are present in the dataset.
This is usually not a major problem as we should have some prior knowledge about the identity of the general cell population,
e.g., we know we've isolated T cells^[
scRNA-seq experiments aren't exactly cheap, either, so we should have some idea about what we're putting into the sequencer.].
Nonetheless, if "absolute" identification of cell types is desired, we need to use cell type annotation methods like `r Biocpkg("SingleR")`.

### Summarizing pairwise effects {#marker-effect-summaries}

As mentioned above, we perform pairwise comparisons between clusters to find differentially expressed genes.
For each gene, we have one effect size of a given type from each pair of clusters;
so in a dataset with $N$ clusters, each cluster will have $N-1$ effect sizes for consideration.
We summarize these effect sizes for each cluster into key statistics such as the mean and median.
This allows us to create a ranking based on one of the summary statistics for a given effect size.

The mean and median are the most obvious and general-purpose summary statistics. 
For cluster $X$, a large mean effect size indicates that the gene is upregulated in $X$ compared to the average of the other groups.
Similarly, a large median effect size indicates that the gene is upregulated in $X$ compared to most (>50%) other clusters.
The median is more robust (or less sensitive, depending on one's perspective) than the mean to large effect sizes in a minority of comparisons, which may or may not be desirable.
In practice, these summaries usually generate similar rankings.

```{r}
previewMarkers(chosen.markers.10x, pre.columns='Symbol', order.by="cohens.d.median")
```

The minimum value is the most stringent summary for identifying upregulated genes.
A large minimum value indicates that the gene is upregulated in $X$ compared to all other clusters.
Ranking on the minimum is a high-risk, high-reward approach;
it can yield a concise set of excellent markers that are unique to $X$,
but can also overlook interesting genes if they are expressed at a similar level in any other cluster. 
The latter effect is not uncommon if the clusters correspond to closely-related cell types.
To give a concrete example, consider a mixed population of CD4^+^-only, CD8^+^-only, double-positive and double-negative T cells.
Neither _Cd4_ or _Cd8_ would be detected as subpopulation-specific markers because each gene is expressed in two subpopulations. 

```{r}
previewMarkers(chosen.markers.10x, pre.columns='Symbol', order.by="cohens.d.min")
```

Another interesting summary statistic is the minimum rank, a.k.a., "min-rank".
The min-rank is the smallest rank of each gene across all pairwise comparisons involving our cluster of interest $X$.
Specifically, genes are ranked within each pairwise comparison based on decreasing effect size, and then the smallest rank across all comparisons is reported for each gene.
A gene with a small min-rank is one of the top upregulated genes in at least one comparison between $X$ and another cluster.
Or in other words: the set of all genes with a min-rank less than or equal to $R$ is equal to the union of the top $R$ genes from all pairwise comparisons for $X$.
This guarantees that our set contains at least $R$ genes that can distinguish our cluster of interest from any other cluster,
which enables a comprehensive determination of a cluster's identity.

```{r}
previewMarkers(chosen.markers.10x, pre.columns='Symbol', order.by="cohens.d.min.rank")

# min.rank <= 5 means represents the union of the top 5 genes from each
# pairwise comparison between our chosen cluster and every other cluster.
chosen.markers.10x$Symbol[chosen.markers.10x$cohens.d.min.rank <= 5]
```

The flexibility to choose between different summary statistics is one of the strengths of our pairwise strategy.
This allows us to explore different rankings of markers depending on our preferences.
For example, the min-rank is a conservative choice as it guarantees separation of our cluster of interest, at the cost of including weaker DE genes in the top set of markers;
while the minimum provides an aggressive ranking that focuses on markers that are uniquely expressed in our cluster (if any exist).
Pairwise comparisons are also robust to differences in the relative number of cells between clusters,
which ensures that a single large cluster does not dominate the calculation of effect sizes for all other clusters.

<!---
We can compare this to a hypotehtical alternative approach where we define markers for each cluster by treating all other cells as a single group. 
This is simpler as it yields a single effect size per gene, but the results will depend on the (potentially unpredictable) composition of subpopulations in the "other" group.
If there is a single large subpopulation that makes up the majority of cells, its expression profile will determine the effect sizes for all other clusters.
If it expresses a gene that is also a marker for your cluster (relative to the other subpopulations besides the dominant one) - too bad, you won't be seeing it in that cluster's ranking.
--->

## Visualizing marker genes 

At this point, we suppose that we ought to create some figures to keep everyone entertained.
We have already demonstrated how we can examine the distribution of expression values with violin plots in Figure \@ref(fig:marker-violins-pbmc).
Another option is to color our $t$-SNE plot according to the log-expression values of a specific marker in each cell (Figure \@ref(fig:marker-tsne-pbmc)).
Any heterogeneity in expression within our cluster might be indicative of internal structure.

```{r marker-tsne-pbmc, fig.wide=TRUE, fig.asp=0.5, fig.cap=paste0("$t$-SNE plot of the cells in the PBMC dataset, colored by the assigned cluster (top) or the log-expression of the top marker gene in cluster ", chosen.cluster, " (bottom).")}
gridExtra::grid.arrange(
    plotReducedDim(sce.nn.10x, "TSNE", colour_by="clusters"),
    plotReducedDim(sce.nn.10x, "TSNE", colour_by=chosen.markers.10x$Symbol[1], swap_rownames="Symbol"),
    ncol=2
)
```

The heatmap is a classic visualization in genomics, and scRNA-seq is no exception (Figure \@ref(fig:marker-heat-pbmc)).
This provides a compact summary of the relative expression of multiple markers across the cell population.
Ideally, each marker should be consistently upregulated in our cluster of interest compared to the rest of the cells in the population.

```{r marker-heat-pbmc, fig.wide=TRUE, fig.cap=paste0("Heatmap of the top markers for cluster ", chosen.cluster, " in the PBMC dataset. Each row represents a gene and each column represents a cell. Each entry is colored by the log-fold change for each cell from the mean log-expression for that gene.")}
plotHeatmap(sce.nn.10x, features=chosen.markers.10x$Symbol[1:10], order_columns_by="clusters", swap_rownames="Symbol", center=TRUE)
```

Another popular visualization is the `r CRANpkg("seurat")`-style "dot plot"^[
Back in my day, a "dot plot" referred to a visualization of a pairwise sequence alignment.
Bet most of you young'uns don't know about that.],
also known as a bubble plot (Figure \@ref(fig:marker-dot-pbmc)).
This is more concise than the heatmap as it uses the size of each dot/bubble to represent the proportion of cells with non-detected expression.

```{r marker-dot-pbmc, fig.wide=TRUE, fig.cap=paste0("Dot plot of the top markers for cluster ", chosen.cluster, " in the PBMC dataset. The size of each point represents the number of cells that express each gene in each cluster, while the color of each point represents the log-fold change between the cluster and the average across all clusters.")}
plotDots(sce.nn.10x, features=chosen.markers.10x$Symbol[1:10], group="clusters", swap_rownames="Symbol", center=TRUE)
```

## Using a log-fold change threshold

The Cohen's $d$ and AUC consider both the magnitude of the difference between clusters as well as the variability within each cluster.
If the variability is low, it is possible for a gene to have a large effect size even if the magnitude of the difference is small.
These genes tend to be uninformative for cell type identification, e.g., ribosomal protein genes.
We would prefer genes with larger log-fold changes between clusters, even if they have higher variability.

To favor the detection of such genes, we can compute the effect sizes relative to a log-fold change threshold.
The definition of Cohen's $d$ is generalized to the standardized difference between the observed log-fold change and the specified threshold.
Similarly, the AUC is redefined as the probability of randomly picking an expression value from one cluster that is greater than a random value from the other cluster plus the threshold.
A large positive Cohen's $d$ and an AUC above 0.5 can only be obtained if the observed log-fold change between clusters is significantly greater than the threshold.
(However, a negative Cohen's $d$ or AUC below 0.5 may not represent downregulation; it may just indicate that the observed log-fold change is less than the specified threshold.)

```{r}
markers.threshold.10x <- scoreMarkers.se(
    sce.nn.10x,
    sce.nn.10x$clusters,
    extra.columns='Symbol',
    more.marker.args=list(threshold=2)
)

chosen.markers.threshold.10x <- markers.threshold.10x[[chosen.cluster]]

# Default ordering by the mean of Cohen's d.
previewMarkers(chosen.markers.threshold.10x, pre.columns="Symbol") 

# Also looking at the order by the mean of the AUCs. 
previewMarkers(chosen.markers.threshold.10x, pre.columns="Symbol", order.by="auc.mean")
```

In general, we only use a threshold if irrelevant genes with low variances are interfering with our interpretation of the clusters.
Weakly expressed genes will often have low log-fold changes due to the pseudo-count shrinkage (Chapter \@ref(normalization)),
and genes that separate closely-related clusters will usually have smaller log-fold changes.
Prematurely using a large threshold will prevent the detection of these potentially interesting genes.

## Handling multiple batches {#marker-batch}

Larger datasets may contain multiple batches with uninteresting differences between batches.
These batch effects can interfere with marker gene detection by 
(i) inflating the variance within each cluster and
(ii) distorting the log-fold changes if the cluster composition varies between batches.
To avoid these issues, we block on the batch of origin for each cell when computing the effect sizes.
Let's demonstrate on a mouse trophoblast dataset [@lun2017assessing] generated across two batches:

```{r}
library(scRNAseq)
sce.tropho <- LunSpikeInData("tropho")
sce.tropho$block <- factor(sce.tropho$block)
table(sce.tropho$block) # plate of origin, which is our batch.

# Computing the QC metrics. For brevity, we'll skip the spike-ins. 
library(scrapper)
is.mito.tropho <- which(any(seqnames(rowRanges(sce.tropho))=="MT"))
sce.qc.tropho <- quickRnaQc.se(sce.tropho, subsets=list(MT=is.mito.tropho), block=sce.tropho$block)
sce.qc.tropho <- sce.tropho[,sce.qc.tropho$keep]

# Computing log-normalized expression values.
sce.norm.tropho <- normalizeRnaCounts.se(sce.qc.tropho, size.factors=sce.qc.tropho$sum, block=sce.qc.tropho$block)

# We now choose the top HVGs.
sce.var.tropho <- chooseRnaHvgs.se(sce.norm.tropho, block=sce.norm.tropho$block)

# Running the PCA on the HVG submatrix.
sce.pca.tropho <- runPca.se(sce.var.tropho, features=rowData(sce.var.tropho)$hvg, block=sce.var.tropho$block)

# Doing some graph-based clustering.
sce.nn.tropho <- runAllNeighborSteps.se(sce.pca.tropho)
```

We set `block=` to instruct `scoreMarkers.se()` to perform the pairwise comparisons separately in each batch.
Specifically, for a comparison between two clusters, we compute one effect size per batch where we only use cells in that batch.
By performing comparisons within each batch, we cancel out any batch effects so that they do not interfere with our effect sizes.
The per-batch effect sizes are then averaged across batches to obtain a single value per comparison,
using a weighted mean that accounts for the number of cells involved in the comparison in each batch.
A similar average across batches is computed for the mean log-expression and proportion of detected cells. 

```{r}
markers.tropho <- scoreMarkers.se(sce.nn.tropho, sce.nn.tropho$clusters, block=sce.nn.tropho$block)
previewMarkers(markers.tropho[["1"]])
```

For simplicity, we do not explicitly penalize genes that behave inconsistently across batches.
This is generally unnecessary as the average favors genes with large effect sizes in the same direction in all batches.
That said, it is theoretically possible for a top marker to have highly variable effect sizes across batches, as long as the average is large.
We can check this by visualizing the expression profile of each gene of interest with respect to the batch (Figure \@ref(fig:markers-violin-batch)).
Ideally, each gene would behave consistently across batches as this is the easiest to interpret -
namely, any conclusions about the biological identity of a cluster will be applicable to all batches.

```{r markers-violin-batch, fig.wide=TRUE, fig.cap=paste0("Distribution of expression values for the top-ranked marker gene (", rownames(markers.tropho[["1"]])[1], ") of cluster 1 in the trophoblast dataset. Distributions are shown for each cluster (x-axis) in each batch (panel).")}
plotExpression(
    sce.nn.tropho,
    features=rownames(markers.tropho[["1"]])[1],
    x="clusters",
    colour_by="clusters",
    other_fields="block"
) +
    facet_grid(~block)
```

Blocking assumes that each pair of clusters is present in at least one batch.
In scenarios where cells from two clusters never co-occur in the same batch,
the associated pairwise comparison will be impossible and is ignored during calculation of summary statistics.
This can be problematic in rare situations where the batches are perfectly confounded with the clusters^[
For bulk RNA-seq, my nickname for this kind of experimental design was "DOA", a.k.a., dead on arrival.]. 

## More uses for the marker scores 

Our discussion above focuses on genes that are upregulated in our cluster of interest, as these are the easiest to interpret and experimentally validate.
However, a cluster may occasionally be defined by downregulation of some genes relative to the rest of the cell population.
In such cases, we can reverse the rankings to see if there is any consistent downregulation compared to other clusters.
Alternatively, we can recognize that any downregulated genes in cluster $X$ should manifest as upregulated genes in other clusters when compared to $X$.
By using a summary like the min-rank, we guarantee that these genes will show up somewhere, i.e., as markers of other clusters.
(Other summaries are less effective as the upregulation only applies to comparisons to $X$ and will not yield a large mean/median effect size.)

```{r}
# Ordering by increasing Cohen's d.
reversed.chosen.markers.10x <- chosen.markers.10x[order(chosen.markers.10x$cohens.d.mean),]
previewMarkers(reversed.chosen.markers.10x, pre.columns="Symbol", post.columns="cohens.d.mean")
```

Occasionally, we are only interested in markers for a subset of the clusters.
Imagine that we have a set of closely-related clusters and we want to identify the genes that distinguish these clusters from each other.
The summary statistics generated from all clusters might not be satisfactory as they will not prioritize genes with weak upregulation between related clusters.
(Except for min-rank, where these genes would at least show up near the top of the ranking.
But they would be surrounded by many irrelevant genes from comparisons to other clusters.)
Instead, we can just compute marker scores from the cells in the selected subset of clusters:

```{r}
# Let's pretend that clusters 1, 2 and 3 are of particular interest and
# we want to find markers between them.
subset.clusters.10x <- sce.nn.10x$clusters %in% c("1", "2", "3")
subset.markers.10x <- scoreMarkers.se(
    sce.nn.10x[,subset.clusters.10x],
    sce.nn.10x$clusters[subset.clusters.10x],
    extra.columns="Symbol"
)

# Now let's have a look at the top markers for cluster 2.
previewMarkers(subset.markers.10x[["2"]], pre.columns="Symbol")
```

For more control on the selected markers, we can filter our data frame on the available statistics.
For example, we might only consider genes as markers if they have detected proportions above 50%,
a mean log-expression greater than 1,
an average difference in the detected proportions above 50%,
and an average log-fold change above 1.
We tend to avoid _a priori_ filtering as it is difficult to choose thresholds that are generally applicable.
Nonetheless, it can be useful to refine the set of markers once we know what we're interested in.

```{r}
filtered.markers.10x <- chosen.markers.10x[
    chosen.markers.10x$detected >= 0.5 &
    chosen.markers.10x$mean >= 1 &
    chosen.markers.10x$delta.detected.mean >= 0.5 &
    chosen.markers.10x$delta.mean.mean >= 1,
]
previewMarkers(filtered.markers.10x, pre.columns="Symbol", post.columns=c("delta.detected.mean"))
```

## Invalidity of $p$-values {#marker-p-value-invalidity}

In the past, we used to report $p$-values along with the effect sizes for the detected markers.
After all, Cohen's $d$ and the AUC are closely related to $t$-tests and the Wilcoxon ranked sum test, respectively.
Unfortunately, the statistical interpretation of $p$-values is compromised when identifying cluster-specific markers.

The first issue is that of "data dredging" (also known as fishing or data snooping) when the DE analysis is performed on the same data used to define the clusters.
We are more likely to get a positive result when we use a dataset to test a hypothesis generated from that data.
Or more simply - clustering will separate cells by expression, so of course we will get low $p$-values when we compare between clusters!
To illustrate, let's simulate i.i.d. normal values, perform $k$-means clustering and test for DE between clusters of cells with Wilcoxon ranked sum tests.
We'd expect a uniform distribution of $p$-values under the null hypothesis, but instead it is skewed towards low values (Figure \@ref(fig:pval-dist)).
This means that we can detect "significant" differences between clusters even in the absence of any real substructure in the data.

```{r pval-dist, fig.cap="Distribution of $p$-values from a DE analysis between two clusters in a simulation with no true subpopulation structure."}
set.seed(0)
y <- matrix(rnorm(1000000), ncol=200)
clusters <- kmeans(t(y), centers=2)$cluster
out <- apply(y, 1, FUN=function(x) {
    wilcox.test(x[clusters==1], x[clusters==2])$p.value
})
hist(out, col="grey80", xlab="p-value", main="")
```

Another problem is that the $p$-value calculation treats counts from the same cluster of cells as replicate observations.
This is not the most relevant level of replication when cells are derived from the same biological sample (i.e., cell culture, animal or patient).
DE analyses that treat cells as replicates fail to properly model the sample-to-sample variability [@lun2017overcoming].
The latter is arguably the more important level of replication as different samples will necessarily be generated if the experiment is to be replicated.
Indeed, the use of cells as replicates only masks the fact that the sample size is actually one if the experiment involving a single biological sample.
By treating cells as replicates, we overstate our sample size and obtain much lower $p$-values than would be appropriate.

In short, the $p$-values for marker genes don't make much sense from a statistical perspective.
We can still use them for ranking, but at that point, we might as well make our life simpler and use the effect sizes directly.
If we really want to determine whether some markers are "real", the best approach is to perform a separate validation experiment with an independent replicate cell population.
A typical strategy is to use different experimental techniques like FACS, FISH, qPCR or IHC to find a subpopulation that expresses the marker(s) of interest.
This confirms that the subpopulation actually exists and is not an artifact of the scRNA-seq protocol or the computational analysis.

## Gene set enrichment

We can summarize the biological functions of our top-ranked marker genes with gene set enrichment analyses.
Here, we extract predefined sets of genes for specific pathways or processes and check if any gene set is overrepresented among our set of top markers.
This reduces some of the hassle of manually examining the annotation for each gene to assign biological meaning to each cluster. 
To illustrate, we'll use the gene ontology (GO)'s biological process (BP) subcategory,
which defines gene sets associated with known biological processes [@ashburner2000gene].

```{r}
library(msigdbr)
go.bp.df <- msigdbr(species="Homo sapiens", collection="C5", subcollection="GO:BP")
go.bp.sets <- split(go.bp.df$ensembl_gene, go.bp.df$gs_name)
```

We use the hypergeometric test to quantify enrichment of each gene set among the top markers for cluster `r chosen.cluster`.
The $p$-value of each gene set is determined by the number of shared genes between each GO set and the top markers (relative to the size of the GO set).
More strongly enriched sets will have lower $p$-values and should be prioritized for interpretation.
Here, we have chosen the top 100 markers but different values can be used depending on how many markers are of interest.

```{r}
# Choosing the top 100 genes with positive Cohen's d values.
top.chosen.10x <- head(rownames(chosen.markers.10x)[chosen.markers.10x$cohens.d.mean > 0], 100)

library(scrapper)
enrich.chosen.10x <- testEnrichment(top.chosen.10x, go.bp.sets, universe=rownames(chosen.markers.10x))

# Prioritizing the most enriched gene sets for examination.
enrich.chosen.10x <- enrich.chosen.10x[order(enrich.chosen.10x$p.value),,drop=FALSE]
head(enrich.chosen.10x)
```

If we need more detail about a particular set, we can examine the behavior of its constituent genes.

```{r}
top.set.10x <- rownames(enrich.chosen.10x)[1]
overlaps.top.set.10x <- intersect(go.bp.sets[[top.set.10x]], top.chosen.10x)
previewMarkers(
    chosen.markers.10x[rownames(chosen.markers.10x) %in% overlaps.top.set.10x,],
    pre.column="Symbol",
    rows=NULL # List all of the top genes in this set.
)
```

Alternatively, we can aggregate the expression profiles of each set's genes into a single per-cell score.
Scores are defined as the column sums of a rank-1 approximation of the submatrix of the log-expression values corresponding to the genes in the set [@bueno2016comprehensive].
This effectively performs a PCA to collapse the submatrix into a single dimension, enriching for the biological signal associated with the set's annotated function.
The resulting scores are primarily useful for visualizing set activity (Figure \@ref(fig:marker-tsne-gsea)).
We tend not to use gene set scores for quantitative analyses as they are difficult to interpret.
Should the score be higher in a cell that weakly upregulates many genes in the set, or a cell that strongly upregulates a few genes in the set?
What if two cells have the same score but express different subsets of genes in the set?
These complications can be minimized by operating on individual genes whenever possible -
for example, instead of testing for differences in gene set scores between subpopulations,
we could examine the distribution of effect sizes for the same comparison across all genes in the set,
which is easier to interpret and more informative.

```{r marker-tsne-gsea, fig.cap=paste0("$t$-SNE plot of the cells in the PBMC dataset, colored by the activity of the `", top.set.10x, "` gene set.")}
top.set.score.10x <- scoreGeneSet.se(sce.nn.10x, go.bp.sets[[top.set.10x]])
plotReducedDim(sce.nn.10x, "TSNE", colour_by=data.frame(Score=top.set.score.10x$scores))
```

Note that many Bioconductor packages implement methods for quantifying gene set enrichment, e.g., `r Biocpkg("fgsea")`, `r Biocpkg("goseq")`, `r Biocpkg("limma")`, to name a few.
We like the hypergeometric test as it is simple and focuses on the top markers, but any function can be used if it can accept a ranking of genes.
In all cases, we would recommend only using the enrichment $p$-values to rank the gene sets, not to make any statements about statistical significance.
Many of these methods will assume that genes are independent under the null hypothesis to compute a $p$-value.
In a biological system with highly coordinated pathways and processes, this is unlikely to be true, potentially inflating the type I error beyond the threshold for significance.

<!---
Focusing on the top markers add some robustness as changes in rank among lower-ranked genes has no effect on the result.
By comparison, other methods care about the entire ranking so if a mild shift in the ranking of mostly-irrelevant genes can affect the p-value.
-->

## Session information {-}

```{r}
sessionInfo()
```
