---
output:
  html_document
bibliography: ref.bib
---

# Multi-condition analysis

```{r setup, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, error=FALSE)
```

## Motivation

Some of our most interesting scRNA-seq datasets consist of multiple samples collected across different experimental conditions.
The idea is to partition cells into putative cell types or states as previously described (Chapters \@ref(clustering), \@ref(batch-correction)),
and then identify changes in gene expression or abundance between conditions for each cell type/state.
This can yield some useful insights on the differences between the conditions, which is a nice change from the descriptive nature of most single-cell analyses.
In fact, a multi-condition analysis represents one of the rare cases where we will actually do some formal hypothesis testing.
So, this chapter is our chance to claw our way back towards some semblance of statistical rigor.

## Differential expression

Differential expression (DE) is an obvious low-hanging fruit when it comes to detecting differences between conditions.
Specifically, our goal is to test for DE between conditions within each cell type/state identified from the single-cell data.
This resolves changes in expression to specific subpopulations, which is more informative than the corresponding bulk RNA-seq analysis^[Though also more expensive.].
To illustrate, let's pull out some pancreas data generated from normal donors and patients with type II diabetes [@segerstolpe2016singlecell]:

```{r}
library(scRNAseq)
sce.seger <- SegerstolpePancreasData()
table(sce.seger$individual, sce.seger$disease)
```

Typically, we would cluster the cells (possibly after batch correction) and then assign some biological interpretation to each cluster.
Happily enough, the authors provided cell type labels so we'll use those directly instead of defining clusters ourselves.

```{r}
sce.seger <- sce.seger[,!is.na(sce.seger$`cell type`)]
table(sce.seger$`cell type`)
```

We compute "pseudo-bulk" expression profiles [@tung2017batch] by summing counts together for all cells with the same combination of cell type and sample.
As their name suggests, these pseudo-bulk profiles are intended to mimic bulk RNA-seq data so that they can be analyzed with existing DE workflows, e.g., `r Biocpkg("edgeR")`, `voom()`.
We use the sum of counts for several reasons:

- Larger counts are more amenable to analysis workflows designed for bulk RNA-seq data.
  Normalization is more straightforward and certain statistical approximations are more accurate,
  e.g., the saddlepoint approximation for quasi-likelihood methods or normality for linear models.
- Collapsing cells into samples reflects the fact that our biological replication occurs at the sample level [@lun2017overcoming].
  Each sample is represented no more than once for each condition, avoiding problems from unmodelled correlations between samples. 
  Supplying the per-cell counts directly to a bulk RNA-seq workflow would imply that each cell is an independent biological replicate,
  which is not true from an experimental perspective.
  (A mixed effects model can handle this variance structure but involves [extra complexity](https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html),
  typically for little benefit - see @crowell2019discovery.)
- Variance across cells within each sample is ignored, provided it does not affect the variance across (replicate) samples.
  This is generally desirable for multi-condition analyses where the primary goal is to find consistent differences between conditions.
  Consider a gene with a strong and consistent change in expression upon treatment but only in a subset of the cells.
  We would happily consider this as a DE gene between conditions, despite the fact that the variance across cells in the treated samples would be inflated.
  By comparison, marker genes should be consistently up- or down-regulated in all cells between subpopulations,
  hence the use of the per-cell expression values in Chapter \@ref(marker-detection).

```{r}
library(scrapper)
pseudo.bulk.seger <- aggregateAcrossCells.se(sce.seger, colData(sce.seger)[,c("individual","cell type")])
pseudo.bulk.seger
colData(pseudo.bulk.seger)[,c("factor.individual", "factor.cell type", "counts")]
assay(pseudo.bulk.seger)[1:10,1:10] # sum of counts for each individual/cell-type combination
```

Once we've generated the pseudo-bulk count matrix, we test for differences between conditions - in this case, disease status - within each cell type.
Any DE analysis method that works with bulk RNA-seq data can be used, provided that we have replicates within each condition.
Here, we'll be using `voom()` from the `r Biocpkg("limma")` package [@law2014voom].
We won't go into too much detail as there is plentiful documentation elsewhere, e.g., in the `limmaUsersGuide()`,
though there are a few pieces of advice that are specific to pseudo-bulk samples:

- Consider removing unreliable pseudo-bulk profiles with very few cells.
  The exact threshold depends on the dataset, the rarity of the cell type, the variance of the assay technology (e.g., UMIs versus reads),
  and whether the DE analysis supports downweighting of low-quality profiles.
  A good rule of thumb seems to be 10 cells [@crowell2019discovery].
- Perform a separate analysis for each cell type instead of cramming all cell types into the same design matrix.
  This protects against differences in the mean-variance relationship across cell types.
  It also ensures that any odd behavior for one cell type does not affect results for the other cell types.
- Get used to higher variances and fewer DE genes compared to actual bulk RNA-seq data.
  The number of cells contributing to each pseudo-bulk profile is often orders of magnitude less than that used in bulk RNA-seq,
  so the latter will be a more precise assay of the population transcriptome.

We test for disease-associated DE genes in beta cells using `voom()` with additional weighting for sample quality.
Perhaps unsurprisingly, _INS_ is one of the top DE genes.
We could then repeat this analysis on all available cell types, though for brevity's sake, we won't show that here. 
Check out other Bioconductor packages like `r Biocpkg("muscat")`, which implement some convenient functions for iterative DE analyses within cell types. 

```{r}
pseudo.beta.seger <- pseudo.bulk.seger[,which(pseudo.bulk.seger$`factor.cell type` == "beta cell")]

# We can have a look at the number of cells contributing to each profile, in
# case we want to remove low-abundance profiles.
pseudo.beta.seger$counts

library(edgeR)
y.beta.seger <- DGEList(assay(pseudo.beta.seger, "sums"), samples=as.data.frame(colData(pseudo.beta.seger)))

keep.beta.seger <- filterByExpr(y.beta.seger, group=y.beta.seger$samples$disease)
y.beta.seger <- y.beta.seger[keep.beta.seger,]
y.beta.seger <- normLibSizes(y.beta.seger)

design.beta.seger <- model.matrix(~disease, y.beta.seger$samples)
v.beta.seger <- voomWithQualityWeights(y.beta.seger, design.beta.seger)
fit.beta.seger <- lmFit(v.beta.seger)
fit.beta.seger <- eBayes(fit.beta.seger, robust=TRUE)

res.beta.seger <- topTable(fit.beta.seger, sort.by="p", n=Inf, coef=2)
head(res.beta.seger)
```

```{r, echo=FALSE}
stopifnot("INS" %in% head(res.beta.seger)$ID)
```

In effect, we treat our scRNA-seq data analysis as a kind of super-powered _in silico_ fluorescence-activated cell sorting (FACS).
Here, experimental isolation of cell types based on a few surface markers is replaced by computational assignment of cell types based on their transcriptomic profiles.
Our pseudo-bulk DE analysis is analogous to a FACS experiment followed by bulk RNA-seq on the isolated populations. 

## Differential abundance

We can also test for differences in cell type abundance between conditions, i.e., differential abundance (DA).
As we all know, immunologists love to create FACS plots showing some change in the percentages between treatments, e.g., Figure 1A of @richard2018tcell.
Now, we can do the same with scRNA-seq data.
For our pancreas dataset, we create a count matrix of the number of cells assigned to each cell type in each sample.
(If we didn't already have annotated cell types, we could instead consider using a tool like `r Biocpkg("miloR")`,
which performs a DA analysis without requiring explicit assignment of each cells to clusters.)

```{r}
ab.count.seger <- countGroupsByBlock(colData(sce.seger)[,"cell type"], colData(sce.seger)$individual)
ab.count.seger <- unclass(ab.count.seger) # get rid of the weird table class.
ab.count.seger
```

We then apply standard DA pipelines to see which cell types are affected by disease.
In particular, testing for DA is bread-and-butter stuff in the microbiome field,
so we'd recommend checking out some of their [best practices](https://microbiome.github.io/OMA/docs/devel/pages/differential_abundance.html).
Right now, though, this book is hard enough to compile without adding extra dependencies,
so we'll just re-use `r Biocpkg("edgeR")`'s statistical machinery to test for differences in the cell abundance matrix [@robinson2010edgeR]^[
With a hammer like `r Biocpkg("edgeR")`, everything kind of looks like a nail.].

```{r}
y.ab.seger <- DGEList(ab.count.seger)
y.ab.seger$samples$disease <- sce.seger$disease[match(colnames(y.ab.seger), sce.seger$individual)]

keep.ab.seger <- filterByExpr(y.ab.seger, group=y.ab.seger$samples$disease)
y.ab.seger <- y.ab.seger[keep.ab.seger,]

# If we don't normalize, our results will be affected by composition bias. But
# if we use TMM normalization, that would assume that most cell types do not
# have any change in their abundance. Hard to tell which one's worse here.

design.ab.seger <- model.matrix(~disease, y.ab.seger$samples)
fit.ab.seger <- glmQLFit(y.ab.seger, design.ab.seger)
res.ab.seger <- glmQLFTest(fit.ab.seger, coef=2)
topTags(res.ab.seger)
```

It's worth noting that DA and DE are two sides of the same coin as they are both inferred from the per-cell expression profiles.
Consider a scRNA-seq experiment involving two biological conditions with several shared cell types.
We focus on a cell type $X$ that is present in both conditions but contains some DE genes between conditions.
This leads to two possible outcomes:

1. The DE between conditions is strong enough to split $X$ into two separate clusters (say, $X_1$ and $X_2$) in expression space.
   This manifests as DA where $X_1$ is enriched in one condition and $X_2$ is enriched in the other condition.
2. The DE between conditions is not sufficient to split $X$ into two separate clusters, 
   e.g., because our batch correction algorithm identifies them as corresponding cell types and merges them together.
   Thus, the differences between conditions manifest as DE within the single cluster corresponding to $X$.

It is difficult to predict whether a difference between conditions will manifest as DE or DA.
For example, we might see DE for coarser clusters but DA for finer clusters.
We'd recommend performing both DE and DA analyses to ensure that we can detect either possibility.

## More thoughts on statistical rigor {#regrets}

In the sections above, we were fortunate enough to use pre-existing cell type annotations from @segerstolpe2016singlecell.
For actual multi-condition analyses, we would first have to assign our own biological identities to our cell subpopulations.
Typically, this involves some kind of correction to merge shared cell types across samples (Chapter \@ref(batch-correction)),
clustering on the corrected data (Chapters \@ref(clustering)), and finally examination of each cluster's marker genes (Chapter \@ref(marker-detection)).
Each cluster is used as a proxy for a cell state/type identity that is common across samples,
serving as the basis for cell type-specific differential expression or abundance across conditions. 

The most obvious concern here is that the hypothesis testing is performed on the same data used to define the subpopulations.
This represents a form of data snooping that complicates the interpretation of the DE/DA $p$-values.
For example, each cluster will consist of cells with similar expression profiles,
which may (i) artificially deflate the variance across replicates but also (ii) understate the differences between conditions in a DE analysis.
In most cases, though, it's probably fine as the process of defining the cell types/states (clustering or otherwise) is blind to the condition label of each cell.
Any arbitrary placement of cell type/state boundaries in high-dimensional expression space should be more-or-less independent of any differences between conditions.
Of course, we can easily imagine exceptions to this rule but these seem slightly pathological^[
For example, if one condition has many more cells than the other, it would dictate the definition of each subpopulation, which could bias the DA analysis.
Or, at very fine clusterings, the distributional assumptions of some DE tools are violated due to underdispersion.]. 

In our opinion, the use of a common clustering is the real Achilles heel of this strategy in terms of statistical rigor.
We fail to capture the uncertainty in the clustering and its biological interpretation, which reduces confidence in the reproducibility of the results.
Say we discover significant DE/DA for a cell type in our dataset.
If an independent party were to repeat our experiment and analysis, would they be able to reach the same conclusion?
More specifically, would they be able to partition an equivalent cluster and assign the same cell type identity?
Weakly separated cell subtypes might not manifest as separate clusters in a new dataset,
or the ranking of markers might change in a manner that causes the analyst to assign a different biological identity.
We wouldn't know - we can't evaluate the reproducibility of our cell type annotations because we only did the clustering and interpretation once.
The same criticism applies to the interpretation of any common manifold, even if no explicit clustering is performed.

To perform multi-condition analyses "more correctly", we need to process each sample independently to capture the variation in interpretation.
Consider a dataset that has multiple replicate samples for each of multiple conditions.
Our analysis strategy would look something like this:

1. Analyze each sample independently, from quality control to identification of cell types/states from the clusters.
   Specifically, biological meaning should be assigned to cell subpopulations without any information from other samples.
   Indeed, if we were being very careful, we would blind and randomize samples across multiple analysts so that variances in human bias are also modelled during manual annotation^[
   Though this is so exceptionally laborious, it's probably not worth doing anything less serious than a clinical trial.].
   Alternatively, we could use automated cell type annotation tools like `r Biocpkg("SingleR")`;
   these do not require any clustering and can be applied to each sample independently, but assume that our cell types of interest exist in the reference annotation.
2. Match corresponding cell types or states across samples.
   This is pretty straightforward if the per-sample cell type/state assignments use a controlled vocabulary, e.g, from the [Cell Ontology](https://cell-ontology.github.io/),
   where the biology of interest is explicitly defined for all samples (and analysts, if more than one person is involved).
   We avoid the use of batch correction to merge cells across samples, which means that we aren't affected by the assumptions and errors of the correction algorithm.
   It's usually at this point that people often start complaining about their favorite cell types/states not showing up consistently across samples.
   But frankly, if the biology is real, it had better be reproducible across your replicates^[
   If you cant't even find your cell type in your own replicates, what chance is there of reproducing it in an independent study?], otherwise it's just wishful thinking.
3. Create a pseudo-bulk or cell abundance count matrix based on the annotated cell types/states from all samples.
   Any variability in the per-sample analysis will manifest as greater variance across replicates in these count matrices.
   For example, if a cell subtype is weakly defined, we may not be able to identify it consistently across replicates, increasing the variance in the cell type abundances.
   Similarly, if a subtype is poorly separated from its relatives, its cluster may occasionally include cells from neighboring subtypes, increasing the variance of the pseudo-bulk profiles.
   The increased variance is important as it properly reflects our uncertainty about the existence of the cell subtype itself.

In practice, this kind of analysis is pretty exhausting, especially for larger studies.
We've only seen this approach used on a handful of occasions over the years because it's just too inconvenient. 
Besides, the incentives for reproducibility don't exist in most scientific environments^[ 
Why should we do more work to introduce more variance and reduce the number of significant hits?
This is antithetical to the raison d'être of single-cell genomics, which is to create publishable results.].
We typically settle on a compromise between convenience and rigor,
where we still use a common clustering from corrected PCs but invest the extra time and resources into independent validation experiments
(see also suggestions in Section \@ref(marker-p-value-invalidity)).
As long as our conclusions can be validated, we can say that our preceding analyses were "exploratory" and give ourselves a pass for any statistical impropriety.

## Ambient contamination in DE analyses

Alright, back to some prosaic concerns.
Ambient contamination is often present in scRNA-seq datasets generated by high-throughput protocols.
During dissociation, cell lysis releases RNA molecules into the ambient solution.
These molecules are captured and sequenced in each cell's reaction chamber, contributing counts to genes that are not otherwise expressed in that cell. 
In most situations, ambient contamination is just a mild nuisance that can be ignored.
It adds a constant baseline expression profile to all cells from the same sample,
but our analysis focuses on the differences between cell types/states, so any addition more-or-less cancels out in the end.

For DE analyses of multi-condition datasets, ambient contamination is more problematic if the ambient expression profile varies across samples. 
After a major experimental perturbation, strong upregulation of a gene in one cell type can contaminate other cell types in the same sample.
Subsequently, DE genes detected for a particular cell type may be driven by differences in the ambient profiles rather than any intrinsic change in gene regulation. 
To illustrate, let's consider the _Tal1_-knockout (KO) chimera data from @pijuansala2019single.
_Tal1_ is a transcription factor that has known roles in erythroid differentiation,
and the aim of the experiment was to determine if blocking of the erythroid lineage in the early mouse embryo diverted cells to other developmental fates.

```{r}
library(MouseGastrulationData)
sce.tal1 <- Tal1ChimeraData()
sce.tal1
```

We perform a pseudo-bulk DE analysis between wild-type (WT) and KO cells labelled as "neural crest".
The strongest DE genes are the hemoglobins, which are downregulated in the KO cells.
This is rather surprising as these cells do not belong to the erythroid lineage and should not express hemoglobins at all. 
The most sober explanation is that the WT samples contain more hemoglobin transcripts in the ambient solution due to leakage from erythrocytes or their precursors. 
(As an aside, it's worth mentioning that the "replicates" in this study are more technical than biological.
This results in some exaggeration of the significance of the effects, as evidenced by the very low $p$-values.
For the purposes of this section, though, we'll just pretend that these are proper replicates.)

```{r}
pseudo.tal1 <- aggregateAcrossCells.se(
    sce.tal1, 
    factor=colData(sce.tal1)[,c("sample", "celltype.mapped")]
)

# Add blocking factor (based on my second-hand knowledge of the study).
pseudo.tal1$block <- factor(pseudo.tal1$sample %% 2 == 0)

# Subset to our neural crest cells.
pseudo.neural <- pseudo.tal1[,pseudo.tal1$celltype.mapped == "Neural crest"]

# Standard voom analysis. 'tomato' is the factor of interest here, as
# tdTomato-marked cells are those in which Tal1 knockout was induced.
y.neural <- DGEList(
    assay(pseudo.neural, "sums"),
    samples=as.data.frame(colData(pseudo.neural)),
    genes=as.data.frame(rowData(pseudo.neural))
)

keep.neural <- filterByExpr(y.neural, group=y.neural$samples$tomato)
y.neural <- y.neural[keep.neural,]
y.neural <- normLibSizes(y.neural)

# Using a log-fold change threshold of 0.5 here, to avoid prioritizing genes
# with very small log-fold changes when the variance is also very low.
design.neural <- model.matrix(~block + tomato, y.neural$samples)
v.neural <- voomWithQualityWeights(y.neural, design.neural)
fit.neural <- lmFit(v.neural)
fit.neural <- treat(fit.neural, robust=TRUE, lfc=0.5)

res.neural <- topTreat(fit.neural, sort.by="none", n=Inf, coef="tomatoTRUE")
head(res.neural[order(res.neural$P.Value),])
```

```{r, echo=FALSE}
# Checking we have the hemoglobins there.
ordered.neural <- res.neural[order(res.neural$P.Value),]
stopifnot(any(grepl("^Hb[ab]-", ordered.neural$SYMBOL[1:10])))

# Flushing out the SCE to allow GC to free up some memory. 
rm(sce.tal1)
```

To quantify the impact of ambient contamination on the interpretation of our DE results,
we obtain an estimate of the ambient expression profile from the unfiltered count matrix for each sample.
We consider all barcodes with total counts at or below 100 to represent empty droplets [@lun2018distinguishing],
and we sum the counts for each gene across these barcodes to obtain an expression vector representing the ambient profile for each sample.

```{r}
# Extracting an ambient profile from the unfiltered count matrix,
# i.e., prior to any cell calling from Cellranger or emptyDrops.
ambient.tal1 <- list()
for (s in pseudo.neural$sample) {
    raw.tal1 <- Tal1ChimeraData(type="raw", samples=s)[[1]]
    mat <- counts(raw.tal1)
    ambient.tal1[[s]] <- SummarizedExperiment(
        list(sums=cbind(rowSums(mat[,colSums(mat) <= 100]))),
        rowData=rowData(raw.tal1)
    ) 
}

ambient.tal1 <- do.call(cbind, ambient.tal1)
ambient.tal1$tomato <- pseudo.neural$tomato
ambient.tal1$block <- pseudo.neural$block
```

Once we have the ambient counts, we incorporate them into the DE analysis to mitigate the impact of contamination.
The standard approach would be to use a "difference of differences" contrast, but this particular dataset is a bit too messy for that. 
Instead, we'll just identify genes that are DE between the WT and KO ambient profiles and remove them from our pseudo-bulk DE list.
The aim is to filter out potentially problematic genes for which DE might be caused by ambient contamination.
This is a simple but effective strategy that gets rid of the hemoglobins from our top DE genes.

```{r}
y.ambient <- DGEList(
    assay(ambient.tal1),
    genes=rowData(ambient.tal1),
    samples=colData(ambient.tal1)
)

# Re-using the same filter that we used for the pseudo-bulk analysis,
# just to make sure we're using the same set of genes.
y.ambient <- y.ambient[keep.neural,]
y.ambient <- normLibSizes(y.ambient)

design.ambient <- model.matrix(~block + tomato, y.ambient$samples)
v.ambient <- voomWithQualityWeights(y.ambient, design.ambient)
fit.ambient <- lmFit(v.ambient)
fit.ambient <- treat(fit.ambient, robust=TRUE, lfc=0.5)
res.ambient <- topTreat(fit.ambient, sort.by="none", n=Inf, coef="tomatoTRUE")

# Decorate our pseudo-bulk result table with some extra ambient statistics.
res.neural.extra <- res.neural
res.neural.extra$ambient.logFC <- res.ambient$logFC
res.neural.extra$ambient.P.Value <- res.ambient$P.Value

# Only keeping genes that are either (i) not DE in the ambient analysis, 
# or (ii) DE in the wrong direction compared to the pseudo-bulk analysis.
filtered.neural <- sign(res.ambient$logFC) != sign(res.neural$logFC) | res.ambient$adj.P.Val >= 0.1
res.neural.extra <- res.neural.extra[filtered.neural,]
head(res.neural.extra[order(res.neural.extra$P.Value),])
```

```{r, echo=FALSE}
ordered.neural.extra <- res.neural.extra[order(res.neural.extra$P.Value),]
stopifnot(all(!grepl("^Hb[ab]-", ordered.neural.extra$SYMBOL[1:100])))
```

Alternatively, we could attempt to remove the ambient contamination from the count matrix itself, e.g., using tools like `cellbender` or `r CRANpkg("SoupX")`.
I don't have much to say on this topic other than to point out that
(i) estimating the contaminating proportion in each single cell is hard and
(ii) removing it while preserving the mean-variance relationship is hard^[
This reminds me of the old problem of input subtraction in ChIP-seq data, which I wrestled with as part of my PhD.
In the end, we just decided to filter out regions with high input coverage rather than attempting subtraction.
It was a lose-lose situation; either the input counts were small and subtraction had no effect,
or the input counts were large and subtraction broke the expected mean-variance relationship in our count data.]. 
The difficulty of the problem motivates some rather complicated models and I prefer a lighter intervention. 

## Session information {-}

```{r}
sessionInfo()
```
